Comparison of Terrain Following and Cut Cell Grids using a Non-Hydrostatic Model

### 2015-09-30 src/tf-cutcell-comparison/tf-cutcell-comparison.tex
remove old gravityWaves plot, add some more description of theta errors

Following \citet{melvin2010}, the domain is \SI{300}{\kilo\meter} wide and \SI{30}{\kilo\meter} high with a grid resolution of \(\Delta x = \SI{500}{\meter}\) and \(\Delta z^\star = \SI{300}{\meter}\).  \TODO{we want to consider much higher resolutions, too} The mountain profile has the same form as equation~\ref{eqn:resting:mountain}.  Tests are performed with mountain heights of $h_0 = \SI{250}{\meter}$ and \(h_0 = \SI{500}{\meter}\).  As in the resting atmosphere test, $a = \SI{5}{\kilo\meter}$ is the mountain half-width and $\lambda = \SI{4}{\kilo\meter}$ is the wavelength.  On the optimised SLEVE grid, $s_1 = \SI{5}{\kilo\meter}$ is the large scale height, $s_2 = \SI{2}{\kilo\meter}$ is the small scale height and the optimal exponent value $n = 1.35$ is used.

@@

No normal flow is imposed at the top and bottom boundaries and the outlet has a zero gradient boundary condition.  For the Exner function of pressure, hydrostatic balance is prescribed on all boundaries.  The simulation is integrated forward by 5 hours with a timestep $\Delta t = \SI{8}{\second}$.

Test results are compared between the BTF, SLEVE and cut cell grids.  Vertical velocities on the BTF, SLEVE and cut cell grids are visually indistinguishable (not shown).  They agree with the high resolution mass-conserving semi-implicit semi-Lagrangian solution from \citet{melvin2010}.

The initial thermal profile is subtracted from the potential temperature field at the end of the integration to reveal the structure of thermal anomalies.  Once again, the results are similar on all three grids, and results are shown on the BTF and cut cell grids in figures~\ref{fig:gw-theta}a and \ref{fig:gw-theta}c respectively.  However, examining more closely the anomalies in the lee of the mountain for the cut cell grid, figure~\ref{fig:gw-theta}d shows that the bottommost layer is anomalously warm and the layer above it is anomalously cold.  This feature is not present on the BTF grid (figure~\ref{fig:gw-theta}b) or the SLEVE grid (not shown).  \TODO{not present?  I think it is, just much smaller amplitude}

@@

Cut cell grids reduced errors in two of the four tests.  First, in the horizontal advection test, tracer accuracy on the cut cell grid was almost as good as accuracy on a regular grid with no mountain.  Second, in the resting atmosphere test, spurious vertical velocities were two orders of magnitude smaller on the cut cell grid compared with the terrain folllowing grids.

@@

@@

\begin{figure*}
	\centering
	\includegraphics{../fig-gravityWaves-theta/fig-gravityWaves-theta.pdf}
%
	\caption{Anomalies in potential temperature in the gravity waves test after 5 hours with a mountain height, \(h_0 = \SI{250}{\meter}\).  The central domain in the lowest \SI{12}{\kilo\meter} is shown on (a) the BTF grid, and (c) the cut cell grid.  The four lowest layers of each grid are shown for (b) BTF, and (d) cut cell grids, using a narrower potential temperature scale.  The results on the SLEVE grid (not shown) are qualitatively identical to results on the BTF grid.  Axes are in units of \si{\meter}.}
	\label{fig:gw-theta}
\end{figure*}

>>>>

Following \citet{melvin2010}, the domain is \SI{300}{\kilo\meter} wide and \SI{30}{\kilo\meter} high.     \TODO{we want to consider much higher resolutions, too} The mountain profile has the same form as equation~\ref{eqn:resting:mountain}, but the gravity waves tests have a mountain height of $h_0 = \SI{250}{\meter}$.  As in the resting atmosphere test, $a = \SI{5}{\kilo\meter}$ is the mountain half-width and $\lambda = \SI{4}{\kilo\meter}$ is the wavelength.  On the optimised SLEVE grid, $s_1 = \SI{5}{\kilo\meter}$ is the large scale height, $s_2 = \SI{2}{\kilo\meter}$ is the small scale height and the optimal exponent value $n = 1.35$ is used.

@@

No normal flow is imposed at the top and bottom boundaries and the outlet has a zero gradient boundary condition.  For the Exner function of pressure, hydrostatic balance is prescribed on all boundaries.

Test results are compared between the BTF, SLEVE and cut cell grids at several resolutions.
The spatial and temporal resolutions tested are shown in table~\ref{tab:gw-resolutions}.  The lowest resolution is the same as that used by \citet{schaer2002}, and higher resolutions preserve the same aspect ratio.
The vertical resolution is chosen to test various configurations of cut cell grid.  At $\Delta z = \SI{300}{\meter}$, the mountain lies entirely within the lowest layer of cells, while at $\Delta z = \SI{250}{\meter}$ and $\Delta z = \SI{125}{\meter}$ the mountain peak is aligned with the interface between layers.  With increasing resolutions up to $\Delta z = \SI{50}{\meter}$, the orography intersects more layers and becomes better resolved.  \TODO{a reviewer (and Terry) suggested these tests, but we haven't found anything notable with any particular cut cell grid.  so I don't see much value in some of this description.  also not sure how much value there is in having a figure showing several cut cell grids (e.g. 300dz, 250dz, 200dz)}
The simulation is integrated forward by 5 hours and the timestep, $\Delta t = 8 \Delta z / \SI{300}{\second}$, is chosen so that it scales linearly with spatial resolution and, following the original test specified by \citet{schaer2002}, $\Delta t = \SI{8}{\second}$ when $\Delta z = \SI{300}{\meter}$.

At $\Delta z = \SI{300}{\meter}$, vertical velocities on the BTF, SLEVE and cut cell grids are visually indistinguishable (not shown).  They agree with the high resolution mass-conserving semi-implicit semi-Lagrangian solution from \citet{melvin2010}.  \TODO{perhaps I should spot check some of the grids at higher resolutions but, given that the theta fields converge overall, I don't see why vertical velocities contours would do anything funky}

The initial thermal profile is subtracted from the potential temperature field at the end of the integration to reveal the structure of thermal anomalies.
Results are similar across all grid types and resolutions, but examining the anomalies in the lee of the mountain errors are found in the lowest few layers: the bottommost layer is anomalously warm and the layer above it is anomalously cold.  \TODO{this isn't true at all resolutions, e.g. 50dz is cold at the bottom}  At all resolutions, the amplitude of the errors is greater on the cut cell grid than the BTF \TODO{or SLEVE?} grid.  As resolution increases, the anomalously cold layers first increase in amplitude, then begin to span multiple layers, which suggests that the errors are controlled by the mountain height. \TODO{that conclusion might be stronger with tests at different mountain heights?  oh no, not more tests! :-P}

@@

Cut cell grids reduced errors in two of the five tests.  First, in the horizontal advection test, tracer accuracy on the cut cell grid was almost as good as accuracy on a regular grid with no mountain.  Second, in the resting atmosphere test, spurious vertical velocities were two orders of magnitude smaller on the cut cell grid compared with the terrain folllowing grids.

@@

\begin{table}
	\caption{Spatial and temporal resolutions used in the gravity waves test.  The lowest resolution has the same parameters as the original test case specified by \citet{schaer2002}.  At higher resolutions, the vertical resolution is chosen, and horizontal and temporal resolutions are calculated to preserve the same aspect ratio.  \TODO{don't need to include all these resolutions, just those need to do suitable analysis/conclusions}}
	\label{tab:gw-resolutions}
%
\centering
\begin{tabular}{S S S}
\hline\hline
{$\Delta z$ (\si{\meter})} & {$\Delta x$ (\si{\meter})} & {$\Delta t$ (\si{\second})} \\
\hline
300	& 500	& 8 \\
250	& 416.7 & 6.667 \\
200	& 333.3 & 5.333 \\
150	& 250	& 4 \\
125	& 208.3 & 3.333 \\
100	& 166.7 & 2.667 \\
75	& 125	& 2 \\
50	& 83.33 & 1.333 \\
\hline
\end{tabular}
\end{table}

@@
